Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2023-01-08 paper.tex
fixed references

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}, but these
approaches are more challenging to interpret and fit to data.
However, new population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to 
African populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models, mostly in Western Africa
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td},
%Plagnol: YRI 
%Here we present a new method for addressing whether archaic human groups contributed to the modern gene pool (called ancient admixture),
%we find strong evidence for ancient admixture in both a European and a West African population
%
%Hammer: Mandenka, Biaka, and San. Title:
%"Genetic evidence for archaic admixture in Africa"
%Abstract:"allow us to infer that contemporary African populations contain a small proportion of genetic material (≈2%) that introgressed ≈35 kya from an archaic population that split from the ancestors of anatomically modern humans ≈700 kya."
%
%Hey: Strongest signal in Yoruba, Baka
% Abstract: "consistent with human/archaic divergence."
%
%Ragsdale: YRI
%abstract: We estimate that an unidentified, deeply diverged population admixed with modern humans within Africa both before and after the split of African and Eurasian populations, contributing 4 − 8% genetic ancestry to individuals in world-wide populations.
%
%Lorente Galdos: Khoesan, Mbuti, W. AFR
%title:"reveals archaic gene flow"
%
%Durvasula: West African
%title: "Recovering signals of ghost archaic introgression"
%Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans." 
but also in southern \citep{Hammer2011-bx,Lorente-Galdos2019-vz}
and central African \citep{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz}
populations.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence \citep[e.g.,][]{Hsieh2016-gk}. However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C). 
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

>>>>

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}, but these
approaches are more challenging to interpret and fit to data.
However, new population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to 
African populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models, mostly in Western Africa
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td},
%Plagnol: YRI 
%Here we present a new method for addressing whether archaic human groups contributed to the modern gene pool (called ancient admixture),
%we find strong evidence for ancient admixture in both a European and a West African population
%
%Hammer: Mandenka, Biaka, and San. Title:
%"Genetic evidence for archaic admixture in Africa"
%Abstract:"allow us to infer that contemporary African populations contain a small proportion of genetic material (≈2%) that introgressed ≈35 kya from an archaic population that split from the ancestors of anatomically modern humans ≈700 kya."
%
%Hey: Strongest signal in Yoruba, Baka
% Abstract: "consistent with human/archaic divergence."
%
%Ragsdale: YRI
%abstract: We estimate that an unidentified, deeply diverged population admixed with modern humans within Africa both before and after the split of African and Eurasian populations, contributing 4 − 8% genetic ancestry to individuals in world-wide populations.
%
%Lorente Galdos: Khoesan, Mbuti, W. AFR
%title:"reveals archaic gene flow"
%
%Durvasula: West African
%title: "Recovering signals of ghost archaic introgression"
%Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans." 
but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} 
and central African \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz,Hey2018-pw}.
populations.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence \citep[e.g.,][]{Hsieh2016-gk}. However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C). 
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

### 2023-01-08 supplement.tex

\begin{figure}[ht!]
    \centering
    \includegraphics{./figures/supp-refit-lls.pdf}
    \caption{
        \textbf{Log-likelihoods of fits of four models to simulated data.} In
        three scenarios, a simple single origin model and our preferred
        continuous migration and merger with stem migration models, we
        simulated 500 replicates of 500 Mb with the same populations and sample
        sizes from our reported best fit models. We then fit the four models
        reported in the main text to each of these simulated datasets and
        compared log-likelihoods. In each case, the simulated model is
        correctly chosen, even under the single origin scenario. We conclude
        our more complex models (with more parameters) are not simply fitting
        statistical noise.
    }
    \label{fig:refit-lls}
\end{figure}

@@

\begin{table}[ht]
\caption{
    \label{tab:supp-single-origin}
    \textbf{Best-fit parameters from the single-origin model.}
    This model includes a single stem population that splits into contemporary human
    groups. Fixed parameters are specified in Table~\ref{tab:fixed-params}.
    Inferred values are scaled to physical units assuming a generation time of
    29 years. This model gave a log-likelihood of -189,434.
}
\centering
\begin{tabular}[t]{rP{8cm}cc}
    \toprule
    Parameter & Description & Value & 95\% CI \\
    \midrule
    $N_e$ & Ancestral effective population size & 9860 & 9600--10090 \\
    $N_{MH}$ & Size of human lineage between Neanderthal and Nama splits & 20850 & 20640--21220 \\
    $N_{Nama_0}$ & Initial Nama size & 10170 & 9950--10250 \\
    $N_{Nama_F}$ & Final Nama size & 222 & 217--226 \\
    $N_{MSL_0}$ & Initial Mende size & 17280 & 16770--17490 \\
    $N_{MSL_F}$ & Final Mende size & 16730 & 16370--17060 \\
    $N_{EA}$ & Size of East African branch & 7130 & 6940--7130 \\
    $N_{Gumuz_F}$ & Final Gumuz size & 3800 & 3690--3900 \\
    $N_{EP}$ & East African agriculturist size & 12990 & 12720--13140 \\
    $N_{GBR_0}$ & Initial British size & 845 & 836--862 \\
    $N_{GBR_F}$ & Final British size & 12140 & 11870--12300 \\
    $N_{Neand}$ & Neanderthal size & 1850 & 1650--1990 \\
    $T_{Nama}$ & Nama split time (years) & 110900 & 107800--112200 \\
    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $2.83\times10^{-5}$ & $2.61\times10^{-5}$--$3.02\times10^{-5}$ \\
    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.98\times10^{-5}$ & $4.82\times10^{-5}$--$5.31\times10^{-5}$ \\
    $m_{MSL-EA}$ & Mende--East Africa migration rate & $18.7\times10^{-5}$ & $18.7\times10^{-5}$--$19.2\times10^{-5}$ \\
    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.38\times10^{-5}$ & $4.13\times10^{-5}$--$4.54\times10^{-5}$ \\
    $m_{EA-EA}$ & Intra-East Africa migration rate & $43.8\times10^{-5}$ & $41.0\times10^{-5}$--$44.2\times10^{-5}$ \\
    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.661 & 0.657--0.671 \\
    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.280 & 0.270--0.286 \\
    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.150 & 0.147--0.154 \\
    \bottomrule
\end{tabular}
\end{table}

>>>>

\begin{figure}[htb]
    \centering
    \includegraphics{./figures/supp-refit-lls.pdf}
    \caption{
        \textbf{Log-likelihoods of fits of four models to simulated data.} In
        three scenarios, a simple single origin model and our preferred
        continuous migration and merger with stem migration models, we
        simulated 500 replicates of 500 Mb with the same populations and sample
        sizes from our reported best fit models. We then fit the four models
        reported in the main text to each of these simulated datasets and
        compared log-likelihoods. In each case, the simulated model is
        correctly chosen, even under the single origin scenario. We conclude
        our more complex models (with more parameters) are not simply fitting
        statistical noise.
    }
    \label{fig:refit-lls}
\end{figure}

@@

\begin{table}[ht]
\caption{
    \label{tab:supp-single-origin}
    \textbf{Best-fit parameters from the single-origin model.}
    This model includes a single stem population that splits into contemporary human
    groups. Fixed parameters are specified in Table~\ref{tab:fixed-params}.
    Inferred values are scaled to physical units assuming a generation time of
    29 years. This model gave a log-likelihood of -189,434.
}
\centering
\begin{tabular}[t]{rP{8cm}cc}
    \toprule
    Parameter & Description & Value & 95\% CI \\
    \midrule
    $N_e$ & Ancestral effective population size & 9910 & 9600--10090 \\
    $N_{MH}$ & Size of human lineage between Neanderthal and Nama splits & 20890 & 20640--21220 \\
    $N_{Nama_0}$ & Initial Nama size & 10170 & 9950--10250 \\
    $N_{Nama_F}$ & Final Nama size & 223 & 217--226 \\
    $N_{MSL_0}$ & Initial Mende size & 17250 & 16770--17490 \\
    $N_{MSL_F}$ & Final Mende size & 16760 & 16370--17060 \\
    $N_{EA}$ & Size of East African branch & 7130 & 6940--7130 \\
    $N_{Gumuz_F}$ & Final Gumuz size & 3830 & 3690--3900 \\
    $N_{EP}$ & East African agriculturist size & 13010 & 12720--13140 \\
    $N_{GBR_0}$ & Initial British size & 843 & 836--862 \\
    $N_{GBR_F}$ & Final British size & 12150 & 11870--12300 \\
    $N_{Neand}$ & Neanderthal size & 1850 & 1650--1990 \\
    $T_{Nama}$ & Nama split time (years) & 110700 & 107800--112200 \\
    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $2.82\times10^{-5}$ & $2.61\times10^{-5}$--$3.02\times10^{-5}$ \\
    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.98\times10^{-5}$ & $4.82\times10^{-5}$--$5.31\times10^{-5}$ \\
    $m_{MSL-EA}$ & Mende--East Africa migration rate & $18.7\times10^{-5}$ & $18.7\times10^{-5}$--$19.2\times10^{-5}$ \\
    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.42\times10^{-5}$ & $4.13\times10^{-5}$--$4.54\times10^{-5}$ \\
    $m_{EA-EA}$ & Intra-East Africa migration rate & $41.4\times10^{-5}$ & $41.0\times10^{-5}$--$44.2\times10^{-5}$ \\
    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.659 & 0.657--0.671 \\
    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.280 & 0.270--0.286 \\
    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.150 & 0.147--0.154 \\
    \bottomrule
\end{tabular}
\end{table}
